An OpenCL-based tridiagonal solver for evaluation of compact finite differences on parallel architectures

### 2015-11-18 thesis/abstract.tex
corrections - round 1

We study the impact of the matrix structure on
the cyclic reduction algorithm
and show that that precomputing
forward reduction coefficients can be
an especially effective strategy.
his is the abstract

>>>>

\emph{Compact} finite difference schemes are widely used
in the direct numerical simulation of fluid flows
for their ability to better resolve the small scales of turbulence.
However, they can be expensive to evaluate and
difficult to parallelize.
In this work, we present an approach for the computation
of compact finite differences
and similar tridiagonal schemes
on graphics processing units (GPUs).
We present a variant of the \emph{cyclic reduction} 
algorithm for solving the
tridiagonal linear systems that arise in such numerical schemes.
We study the impact of the matrix structure on
the cyclic reduction algorithm
and show that precomputing
forward reduction coefficients can be
especially effective for obtaining good performance.
Our tridiagonal solver is able to outperform
the NVIDIA CUSPARSE
and the multithreaded Intel MKL tridiagonal solvers
on GPU and CPU respectively.
In addition,
we present a parallelization strategy for GPU-accelerated clusters,
and show scalabality
of a 3-D compact finite difference application
for up to 64 GPUs on Clemson's Palmetto cluster.

### 2015-11-18 thesis/acknowledgments.tex
corrections - round 1

I owe my deepest thanks
to my advisor, Dr. Richard S. Miller,
for hist immense guidance and support;
and the members of my committee,
Dr. Gang Li and Dr. Lonny Thompson
for their valuable time.
I would like to acknowledge
Dr. Daniel Livescu at Los Alamos National Lab
for introducing us to this interesting problem,
for his helpful comments,
and for generous funding.
I also thank
Dr. Melissa Smith and Karan Sapra
for their helpful guidance,
and for organizing an excellent course on GPU computing.
I owe a great deal of gratitude to
Dr. Christopher Cox at Clemson University
for his guidance on a range of topics
and for many interesting conversations.
I would like to thank the
staff of the Cyberinfrastructure Technology Integration,
and especially
Dr. Galen Collier and Dr. Marcin Ziolkowski
for their valuable comments and support.
I owe thanks to Clemson University for the generous
allocation of compute time on the Palmetto cluster.
I acknowledge
Dr. K. N. Seetharamu and Dr. V. Krishna,
my advisors at PES Institute of Technology
for inspiring in me a passion for
scientific research.
I owe a special thanks to the
Software Carpentry Foundation
and its members,
for helping me be a
a more productive programmer,
a more deliberate learner,
a more effective teacher,
and a better person.

>>>>

I owe my deepest thanks
to my advisor, Dr. Richard S. Miller,
for hist immense guidance and support;
and the members of my committee,
Dr. Gang Li and Dr. Lonny Thompson
for their valuable time.
I would like to acknowledge
Dr. Daniel Livescu at Los Alamos National Lab
for introducing us to this interesting problem,
for his helpful comments,
and for generous funding.
I also thank
Dr. Melissa Smith and Karan Sapra
for their helpful guidance,
and for organizing an excellent course on GPU computing.
I owe a great deal of gratitude to
Dr. Christopher Cox at Clemson University
for his guidance on a range of topics
and for many interesting conversations.
I would like to thank the
staff of the Cyberinfrastructure Technology Integration,
and especially
Dr. Galen Collier and Dr. Marcin Ziolkowski
for their valuable comments and support.
I owe thanks to Clemson University for the generous
allocation of compute time on the Palmetto cluster.
I acknowledge
Dr. K. N. Seetharamu and Dr. V. Krishna,
my advisors at PES Institute of Technology
for inspiring in me a passion for
scientific research.
I am indebted to Dr. Anush Krishnan
for introducing me to the fascinating field of
scientific computing.
I owe a special thanks to the
Software Carpentry Foundation
and its members,
for helping me be
a more productive programmer,
a more deliberate learner,
a more effective teacher,
and a better person.

### 2015-11-18 thesis/introduction-and-background.tex
corrections - round 1

In the direct numerical simulation (DNS) of fluid systems,
the aim is to resolve the smallest features of the flow
without resorting to any modeling of the turbulence.
This involves the use of
extremely fine computational grids to
discretize the flow domain,
and some numerical method for solving the
flow equations on this grid.
By far the most popular numerical approach for
solving the flow equations
is the \emph{finite difference method}.
The finite difference method approximates the
spatial and temporal derivatives appearing in the
partial differential equations that describe the flow
using finite difference schemes.
\emph{Compact finite difference schemes} are a class
of finite difference schemes that have found widespread
adoption in DNS codes
for their high order of accuracy.
The evaluation of compact finite differences
requires the repeated solution of \emph{banded} linear systems,
making them fairly complex and expensive computationally.

@@

The objective of this work is to develop an
approach for evaluating
compact finite differences
on multiple GPUs in a distributed system.
This is of interest because
the evaluation of spatial derivatives using
compact finite difference schemes is one of the most
expensive tasks in DNS.
But perhaps more significantly,
it encompasses several computational patterns
such as pointwise updates, stencil evaluations,
and solutions of distributed tridiagonal systems.
Efficient implementation of these computational patterns
on GPUs is of interest in other areas of CFD,
and scientific computation in general.

@@

Compact schemes express the derivative at a point $i$
in terms of
function values \emph{and derivatives}
at neighbouring points,
i.e., the derivative is expressed \emph{implicitly}.
For example,
if $f_i$ represents the value of
a uniformly sampled function evaluated at the $i$th sample point,
the first derivative $f^{\prime}_i$ can be approximated from
a relation of the form:
%
\begin{equation}
\begin{split}
    \alpha(f^{\prime}_{i-1} + f^{\prime}_{i+1}) + \
        f^{\prime}_i
    \beta(f^{\prime}_{i-2} + f^{\prime}_{i+2}) + \
    \hdots
    = 
    a\frac{f_{i+1} - f_{i-1}}{dx} + \
    b\frac{f_{i+2} - f_{i-2}}{dx} + \\
    c\frac{f_{i+3} - f_{i-3}}{dx} + \
    \hdots
\end{split}
\label{eqn:general-compact}
\end{equation}
%
where $\alpha$, $\beta$, $a$, $b$, $c$, etc.,
are parameters that must satisfy certain constraints ~\cite{lele1992compact}.
We note that for equations of the form in Eq. \ref{eqn:general-compact},
the derivative at any points $i$ cannot be computed explicitly.
Instead, we must write similar equations for \emph{all} points $i$
in the range.
This results in a system of linear equations
with unknowns \{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
which may be represented by the matrix system $A\bm{x}=\bm{d}$,
where $\bm{x}$ is the vector
\{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
$\bm{d}$ is a vector of right hand sides [Eq.\ref{eqn:general-compact}],
and $A$ is, in general, a \emph{banded} matrix.

@@

\begin{figure}
\begin{center}
\includegraphics[height=250pt]{fig/modified-wavenumbers.eps}
\end{center}
\caption{Modified wavenumbers for three different schemes
    with the same stencil width.
    The compact schemes better estimate the derivative
    for higher wavenumbers.}
\label{fig:modified-wavenumbers}
\end{figure}

@@

From \ref{fig:modified-wavenumbers},
it is clear that the compact schemes are better
able to compute derivatives for larger wavenumbers
for a given stencil width.
This makes them the superior choice for DNS,
in which the flow quantities
exhibit spatial variation associated with high wavenumbers.

@@

One of the advantages of the compact finite difference approach
is that it accommodates \emph{non-periodic} boundary conditions.
This is in contrast to other methods used in DNS,
such as the \emph{spectral} methods.
We note that the Eq. (\ref{eqn:general-compact})
cannot be applied near the boundary points.
At the boundaries,
\emph{non-centered} or \emph{one-sided}
finite difference approximations are required.
Some considerations are made in choosing
these approximations:
firstly, the bandwidth of the resulting banded matrix
must be preserved.
Secondly,
the width of the boundary stencils
must be lower than the interior stencils,
as higher order boundary stencils are unstable
~\cite{kennedy1994several}.
In general, boundary equations for the first derivative
are of the following form ~\cite{lele1992compact}:

\begin{equation}
    f_1^{\prime} + \alpha f_2^{\prime} = \
        \frac{1}{h}(af_1 + bf_2 + cf_3 + df_4)
\label{eqn:boundary-compact}
\end{equation}

@@

As mentioned in Sec. \ref{subsec:gen-form-compact},
compact finite difference schemes lead to
\emph{banded} linear systems.
The simplest classes of banded matrix include
diagonal, tridiagonal and pentadiagonal matrices.
When $\alpha$ = $\beta$ = $\hdots$ = 0,
$A$ is a $\emph{diagonal}$ matrix.
In this case, the
scheme is \emph{explicit}
and the derivatives are straightforward to evaluate:
this simply involves the
application of the right-hand side stencil at each point.
When $\alpha \neq 0, \beta = \hdots$ = 0,
$A$ is \emph{tridiagonal}.
The evaluation of the derivatives requires
the solution of the resulting tridiagonal system.
When $\alpha \neq 0$ and $\beta \neq  0$,
pentadiagonal and higher-bandwidth systems arise.
These can be more expensive and difficult to evaluate
than tridiagonal systems
using a direct method.

This work will focus on
compact finite difference schemes that lead
to tridiagonal systems.
An example of such a scheme is obtained
by substituting 
$\beta = 0$, $\alpha = \frac{1}{4}$ and $a = \frac{3}{4}$
in Eq. \ref{eqn:general-compact}
(all other coefficients are set to zero).
This leads to a fourth-order accurate
tridiagonal compact finite difference scheme,
known also as the Pad\'{e} scheme.
At the boundaries,
the following \emph{third-order} accurate
approximations are used:

@@

      f^{\prime}_2 \\
      f^{\prime}_3 \\
      \vdots \\
      \vdots \\
      \vdots \\
      \vdots \\
      f^{\prime}_{n-1} \\
      f^{\prime}_n
   \end{bmatrix}
 =
 \begin{bmatrix}
     \frac{-5f_1 + 4f_2 + f_3}{2}\\
     \frac{3(f_{3} - f_{1})}{4}\\
     \frac{3(f_{4} - f_{2})}{4}\\
     \vdots\\
     \vdots\\
     \vdots\\
     \vdots\\
     \frac{3(f_{n} - f_{n-2})}{4}\\
     \frac{5f_{n} - 4f_{n-1} - f_{n-2}}{2}
  \end{bmatrix}
\end{equation}

@@

CUDA is the name for the parallel computing platform
developed by NVIDIA,
as well as the application programming interface
for programming their GPUs.
CUDA allows general-purpose applications to be
more easily programmed for the GPU.
Two essential features of the CUDA programming modl
are \emph{kernels} and \emph{thread organization}.

@@

From the hardware perspective,
an NVIDIA GPU may be viewed primarily as
a collection of so-called Streaming Microprocessors (SMs).
When a kernel is launched with specified
grid size (number of blocks)
and block size (threads per block),
each block is assigned to an SM.
Threads within a thread block can execute concurrently on the SM,
and an SM can execute several thread blocks concurrently.
It is imperative that blocks are able to run
independently, in series or in parallel.
This feature gives CUDA programs
their scalability \ref{fig:gpu-scaling}.
The same program runs faster on a GPU with a larger number
of SMs, simply because more blocks may run concurrently.
The number of thread blocks that an
SM can execute concurrently
is limited by a number of factors,
and maximizing this number is often
key to obtaining good performance.
Another key consideration in the implementation
of algorithms on GPUs is the
available \emph{memory hierarchy}.
Threads executing a kernel can read and write
to several different memory spaces,
each with different capacity, latency and bandwidth.

@@

Threads within a block have access to a
common, fast \emph{shared memory}.
All threads within a block
can read and write to the block's shared memory,
but they may \emph{not} access
shared memory of another block.
Thread access to shared memory can be expected to be
much faster than access to global memory.
Data that may be repeatedly used in a kernel
are good candidates for placement in shared memory
The contents of shared memory are managed by the kernel code,
and for this reason,
shared memory is often viewed as explicitly managed cache.
Data in shared memory does\emph{not} persist after
kernel execution.

@@

\begin{itemize}
    \item In any application, the data transfers between
        the host and device must be minimized.
        Ideally,
        data is read into the device from the host \emph{once}
        and from the device to the host \emph{once}.
    \item Thread access to global memory must \emph{coalesced},
        i.e., successive threads must read successive locations
        in global memory.
    \item Shared memory access must be free of bank conflicts
        as much as possible---in general, this means avoiding
        strided shared memory access.
    \item The amount of shared memory and registers
        allocated for each block is kept minimum.
\end{itemize}

@@

\begin{figure}
\begin{center}
\includegraphics[height=200pt]{img/gpu-mapping.eps}
\end{center}
\caption{Mapping work to blocks and threads:
systems are mapped to blocks and
indices are mapped to individual threads.}
\label{fig:gpu-mapping}
\end{figure}

In the GPU implementation of cyclic reduction,
blocks are assigned to tridiagonal systems
(when solving multiple systems),
and threads within a block 
are assigned to equations, or \emph{indices}
(Fig. \ref{fig:gpu-mapping}).
In this way, several grid lines are solved concurrently
by the GPU.
During the forward reduction phase,

@@

the threads assigned to each even index $i$
compute the coefficients
and right hand side for the reduced tridiagonal system
$a_i^\prime$, $b_i^\prime$ and $c_i^\prime$
and $d_i^\prime$.
In practice, the coefficients and right hand side
are updated \emph{in-place}.
Figure \ref{fig:forward-reduction-step} shows the updates
to the coefficient array $\bm{b}$
in the first forward reduction step,
a similar pattern is seen for the arrays
$\bm{a}$, $\bm{c}$ and $\bm{d}$.
In each step of forward reduction,
a thread accesses values from the
coefficient arrays and right hand side
in a \emph{strided} fashion.
At every subsequent step,
this stride is doubled, while the number of active threads is halved
\ref{fig:cyclic-reduction}.
In the backward substitution phase,
the strides are halved at each step,
while the number of active threads is doubled.

Thus, the following issues are encountered
while implementing cyclic reduction for GPUs:

@@

Despite these issues,
cyclic reduction remains an attractive algorithm
for GPUs,
for its low algorithmic complexity,
and lower work per step compared to PCR.
Much work has been done on
optimizing cyclic reduction performance on GPUs,
and addressing the issues discussed.
Zhang et al. \cite{Zhang2010FTS} propose a
\emph{hybrid} solver
that uses both cyclic reduction and parallel cyclic reduction
to reduce the number and severity of bank conflicts,
and also to have better thread activity overall.
G{\"o}dekke et al. \cite{GoSt11CR}
use a method of separately storing
the even and odd indexed equations
to arrive at a bank-conflict free solver
at the cost of additional shared memory usage.
Davidson et al. \cite{davidson2011register}
describe the method of
\emph{register packing}---performing more computations
on \emph{registers}, rather than shared memory---as
a means to reduce shared memory usage in cyclic reduction.
Esfahanian et. al, \cite{esfahanian2014efficient}
avoid shared memory (and associated bank conflicts entirely)
using a data rearrangement scheme to improve global memory access.
Our approach takes a different route,
and is focussed on \emph{near-Toeplitz} systems---exploiting
the relatively simple matrix structure
to reduce the number of computations and memory accesses
performed by each thread
at every cyclic reduction step.

>>>>

In the direct numerical simulation (DNS) of fluid systems,
the aim is to resolve the smallest features of the flow
without resorting to any modeling of the turbulence.
This involves the use of
extremely fine computational grids to
discretize the flow domain,
and some numerical method for solving the
flow equations on this grid.
Among the most popular numerical approaches for
solving the flow equations
is the \emph{finite difference method}.
The finite difference method approximates the
spatial and temporal derivatives appearing in the
partial differential equations that describe the flow
using finite difference schemes.
\emph{Compact finite difference schemes} are a class
of finite difference schemes that have found widespread
adoption in DNS codes
for their high order of accuracy and small stencil widths.
The evaluation of compact finite differences
requires the repeated solution of \emph{banded} linear systems,
making them fairly complex and expensive computationally.

@@

The objective of this work is to develop an
approach for evaluating
compact finite differences---and
other numerical schemes leading to tridiagonal systems---on
multiple GPUs in a distributed system.
This is of interest because
the evaluation of spatial derivatives using
compact finite difference schemes is one of the most
expensive tasks in DNS.
But perhaps more significantly,
it encompasses several computational patterns
such as pointwise updates, stencil evaluations,
and solutions of distributed tridiagonal systems.
Efficient implementation of these computational patterns
on GPUs is of interest in other areas of CFD,
and scientific computation in general.

@@

Compact schemes express the derivative at a point $i$
in terms of
function values \emph{and derivatives}
at neighbouring points,
i.e., the derivative is expressed \emph{implicitly}.
For example,
if $f_i$ represents the value of
a uniformly sampled function evaluated at the $i$th sample point,
the first derivative $f^{\prime}_i$ can be approximated from
a relation of the form:
%
\begin{equation}
\begin{split}
    f_i^{\prime} + \
    \alpha(f^{\prime}_{i-1} + f^{\prime}_{i+1}) + \
    \beta(f^{\prime}_{i-2} + f^{\prime}_{i+2}) + \
    \hdots
    = 
    a\frac{f_{i+1} - f_{i-1}}{dx} + \
    b\frac{f_{i+2} - f_{i-2}}{dx} + \\
    c\frac{f_{i+3} - f_{i-3}}{dx} + \
    \hdots
\end{split}
\label{eqn:general-compact}
\end{equation}
%
where $\alpha$, $\beta$, $a$, $b$, $c$, etc.,
are parameters that must satisfy certain constraints ~\cite{kennedy1994several,lele1992compact},
and $dx$ is the space between two grid points.
We note that for equations of the form in Eq. (\ref{eqn:general-compact}),
the derivative at any points $i$ cannot be computed explicitly.
Instead, we must write similar equations for \emph{all} points $i$
in the range.
This results in a system of linear equations
with unknowns \{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
which may be represented by the matrix system $A\bm{x}=\bm{d}$,
where $\bm{x}$ is the vector
\{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
$\bm{d}$ is a vector of right hand sides [Eq. (\ref{eqn:general-compact})],
and $A$ is, in general, a \emph{banded} matrix.

@@

\begin{figure}
\begin{center}
\includegraphics[height=250pt]{fig/modified-wavenumbers.eps}
\end{center}
\caption{Modified wavenumbers for different finite difference schemes.
    $k^\prime$ is the wavenumber of the approximated derivative,
    while $k$ is wavenumber of the exact derivative.
    $h$ represents the spacing between two grid points.
    The compact schemes better estimate the derivative
    for higher wavenumbers.}
\label{fig:modified-wavenumbers}
\end{figure}

@@

From Fig. \ref{fig:modified-wavenumbers},
it is clear that the compact schemes are better
able to compute derivatives for larger wavenumbers
for a given stencil width.
This makes them the superior choice for DNS,
in which the flow quantities
exhibit spatial variation associated with high wavenumbers.

@@

One of the advantages of the compact finite difference approach
is that it accommodates \emph{non-periodic} boundary conditions.
This is in contrast to other methods used in DNS,
such as \emph{spectral} methods.
We note that the Eq. (\ref{eqn:general-compact})
cannot be applied near the boundary points.
At the boundaries,
\emph{non-centered} or \emph{one-sided}
finite difference approximations are required.
Some considerations are made in choosing
these approximations:
firstly, the bandwidth of the resulting banded matrix
must be preserved.
Secondly,
the width of the boundary stencils
must be lower than the interior stencils,
as higher order boundary stencils are unstable
~\cite{kennedy1994several}.
In general, boundary equations for the first derivative
are of the following form ~\cite{lele1992compact}:

\begin{equation}
    f_1^{\prime} + \alpha^{\prime}f_2^{\prime} = \
        \frac{1}{dx}(a^{\prime}f_1 + b^{\prime}f_2 + c^{\prime}f_3 + d^{\prime}f_4)
\label{eqn:boundary-compact}
\end{equation}

@@

As mentioned in Sec. \ref{subsec:gen-form-compact},
compact finite difference schemes lead to
\emph{banded} linear systems.
The simplest classes of banded matrix include
diagonal, tridiagonal and pentadiagonal matrices.
When $\alpha$ = $\beta$ = $\hdots$ = 0,
$A$ is a $\emph{diagonal}$ matrix.
In this case, the
scheme is \emph{explicit}
and the derivatives are straightforward to evaluate:
this simply involves the
application of the right-hand side stencil at each point.
When $\alpha \neq 0, \beta = \hdots$ = 0,
$A$ is tridiagonal.
The evaluation of the derivatives requires
the solution of the resulting tridiagonal system.
When $\alpha \neq 0$ and $\beta \neq  0$,
pentadiagonal systems arise.
These can be more expensive and difficult to evaluate
than tridiagonal systems
using a direct method.

This work will focus on
compact finite difference schemes that lead
to tridiagonal systems.
An example of such a scheme is obtained
by substituting 
$\beta = 0$, $\alpha = \frac{1}{4}$ and $a = \frac{3}{4}$
in Eq. (\ref{eqn:general-compact})
(all other coefficients are set to zero).
This leads to a fourth-order accurate
tridiagonal compact finite difference scheme,
known also as the Pad\'{e} scheme.
At the boundaries,
the following \emph{third-order} accurate
approximations are used:

@@

      f^{\prime}_2 \\
      f^{\prime}_3 \\
      \vdots \\
      \vdots \\
      \vdots \\
      \vdots \\
      f^{\prime}_{n-1} \\
      f^{\prime}_n
   \end{bmatrix}
 =
 \begin{bmatrix}
     \frac{-5f_1 + 4f_2 + f_3}{2dx}\\
     \frac{3(f_{3} - f_{1})}{4dx}\\
     \frac{3(f_{4} - f_{2})}{4dx}\\
     \vdots\\
     \vdots\\
     \vdots\\
     \vdots\\
     \frac{3(f_{n} - f_{n-2})}{4dx}\\
     \frac{5f_{n} - 4f_{n-1} - f_{n-2}}{2dx}
  \end{bmatrix}
\end{equation}

@@

CUDA is the name for the parallel computing platform
developed by NVIDIA,
as well as the application programming interface
for programming their GPUs.
CUDA allows general-purpose applications to be
more easily programmed for the GPU.
Two essential features of the CUDA programming model
are \emph{kernels} and \emph{thread organization}.

@@

From the hardware perspective,
an NVIDIA GPU may be viewed primarily as
a collection of so-called Streaming Microprocessors (SMs).
When a kernel is launched with specified
grid size (number of blocks)
and block size (threads per block),
each block is assigned to an SM.
Threads within a thread block can execute concurrently on the SM,
and an SM can execute several thread blocks concurrently.
It is imperative that blocks are able to run
independently, in series or in parallel.
This feature gives CUDA programs
their scalability (Fig. \ref{fig:gpu-scaling}).
The same program runs faster on a GPU with a larger number
of SMs, simply because more blocks may run concurrently.
The number of thread blocks that an
SM can execute concurrently
is limited by a number of factors,
and maximizing this number is often
key to obtaining good performance.
Another key consideration in the implementation
of algorithms on GPUs is the
available \emph{memory hierarchy}.
Threads executing a kernel can read and write
to several different memory spaces,
each with different capacity, latency and bandwidth.

@@

Threads within a block have access to a
common, fast \emph{shared memory}.
All threads within a block
can read and write to the block's shared memory,
but they may \emph{not} access
shared memory of another block.
Thread access to shared memory can be expected to be
much faster than access to global memory.
Data that may be repeatedly used in a kernel
are good candidates for placement in shared memory
The contents of shared memory are managed by the kernel code,
and for this reason,
shared memory is often viewed as explicitly managed cache.
Data in shared memory does \emph{not} persist after
kernel execution.

@@

\begin{itemize}
    \item In any application, the data transfers between
        the host and device must be minimized.
        Ideally,
        data is read into the device from the host \emph{once}
        and from the device to the host \emph{once}.
    \item Thread access to global memory must be \emph{coalesced},
        i.e., successive threads must read successive locations
        in global memory.
    \item Shared memory access must be free of bank conflicts
        as much as possible---in general, this means avoiding
        strided shared memory access.
    \item The amount of shared memory and registers
        allocated for each block is kept minimum.
\end{itemize}

@@

\begin{figure}
\begin{center}
\includegraphics[height=200pt]{img/gpu-mapping.eps}
\end{center}
\caption{Mapping work to blocks and threads:
systems are mapped to blocks and
indices are mapped to individual threads.}
\label{fig:gpu-mapping}
\end{figure}
%
In the GPU implementation of cyclic reduction,
blocks are assigned to tridiagonal systems
(when solving multiple systems),
and threads within a block 
are assigned to equations, or \emph{indices}
(Fig. \ref{fig:gpu-mapping}).
In this way, several grid lines are solved concurrently
by the GPU.
During the forward reduction phase,
the threads assigned to each even index $i$
compute the coefficients
and right hand side for the reduced tridiagonal system
$a_i^\prime$, $b_i^\prime$ and $c_i^\prime$
and $d_i^\prime$.
In practice, the coefficients and right hand side
are updated \emph{in-place}.
Figure \ref{fig:forward-reduction-step} shows the updates
to the coefficient array $\bm{b}$
in the first forward reduction step,
a similar pattern is seen for the arrays
$\bm{a}$, $\bm{c}$ and $\bm{d}$.
In each step of forward reduction,
a thread accesses values from the
coefficient arrays and right hand side
in a \emph{strided} fashion.
At every subsequent step,
this stride is doubled, while the number of active threads is halved
\ref{fig:cyclic-reduction}.
In the backward substitution phase,
the strides are halved at each step,
while the number of active threads is doubled.

@@

Several issues are encountered in the GPU implementation:

@@

Despite these issues,
cyclic reduction remains an attractive algorithm
for GPUs,
for its low algorithmic complexity,
and lower work per step compared to PCR.
Much work has been done on
addressing these problems and
optimizing cyclic reduction performance on GPUs.
Zhang et al. \cite{Zhang2010FTS} propose a
\emph{hybrid} solver
that uses both cyclic reduction and parallel cyclic reduction
to reduce the number and severity of bank conflicts,
and also to have better thread activity overall.
G{\"o}dekke et al. \cite{GoSt11CR}
use a method of separately storing
the even and odd indexed equations
to arrive at a bank-conflict free solver
at the cost of additional shared memory usage.
Davidson et al. \cite{davidson2011register}
describe the method of
\emph{register packing}---performing more computations
on \emph{registers}, rather than shared memory---as
a means to reduce shared memory usage in cyclic reduction.
Esfahanian et al. \cite{esfahanian2014efficient}
avoid shared memory (and associated bank conflicts entirely)
using a data rearrangement scheme to improve global memory access.
Our approach takes a different route,
and is focussed on exploiting the specific matrix structure
to reduce the number of computations and memory accesses
performed by each thread
at every cyclic reduction step.

### 2015-11-18 thesis/proposed-tridiagonal-algorithm.tex
corrections - round 1

\begin{equation} \label{eqn:toeplitz-matrix}
A = 
\begin{pmatrix}
     b_1 & c_1  \\
     a_0 & b_0  &  c_0  \\
         & a_0  &  b_0 &  c_0  \\
         &      &  a_0 &  b_0 &  c    \\
         &      &      &      &  \ddots \\
         &      &      &      &     &  \ddots  \\
         &      &      &      &     &  a_n  &  b_n
\end{pmatrix}
\end{equation}
%
where, in general
\begin{align*}
    a_0 &\neq a_n  \\
    b_1 &\neq b_0 \neq b_n \\
    c_1 &\neq c_0 
\end{align*}
%
We refer to matrices with this specific structure as
\emph{near-Toeplitz tridiagonal matrices},
and the corresponding linear systems as
\emph{near-Toeplitz tridiagonal systems}.
These matrices appear in a wide range of applications
~\cite{sun1995application}
such as alternating direction implicit methods,
line relaxation methods,
and numerical solutions to one-dimensional differential equations.
We describe an efficient solver for solving such
near-Toeplitz tridiagonal systems efficiently on GPUs.
\subsection{Forward reduction}

@@

\begin{figure}
\begin{center}
\includegraphics[height=200pt]{img/cyclic-reduction-precomputing.eps}
\end{center}
\caption{Maximum storage required for forward reduction coefficients
at all steps $m=1, 2, ... log_2(n)-1$}
\label{fig:cyclic-reduction-precomputing}
\end{figure}

>>>>

\begin{equation} \label{eqn:toeplitz-matrix}
A = 
\begin{pmatrix}
     b_1 & c_1  \\
     a_0 & b_0  &  c_0  \\
         & a_0  &  b_0 &  c_0  \\
         &      &  a_0 &  b_0 &  c    \\
         &      &      &      &  \ddots \\
         &      &      &      &     &  \ddots  \\
         &      &      &      &     &  a_n  &  b_n
\end{pmatrix}
\end{equation}
%
where, in general
\begin{align*}
    a_0 &\neq a_n  \\
    b_1 &\neq b_0 \neq b_n \\
    c_1 &\neq c_0 
\end{align*}
%
We refer to matrices with this specific structure as
\emph{near-Toeplitz tridiagonal matrices},
and the corresponding linear systems as
\emph{near-Toeplitz tridiagonal systems}.
These matrices appear in a wide range of applications
~\cite{sun1995application}
such as alternating direction implicit methods,
line relaxation methods,
and numerical solutions to one-dimensional differential equations.
We describe an approach for solving such
near-Toeplitz tridiagonal systems efficiently on GPUs.

Below, we present the effect of the matrix structure
[Eq. (\ref{eqn:toeplitz-matrix})] on the
forward reduction and backward substitution phases of
the cyclic reduction algorithm.

\subsection{Forward reduction}

@@

\begin{figure}
\begin{center}
\includegraphics[height=250pt]{img/cyclic-reduction-precomputing.eps}
\end{center}
\caption{Maximum storage required for forward reduction coefficients
at all steps $m=1, 2, ... log_2(n)-1$}
\label{fig:cyclic-reduction-precomputing}
\end{figure}
